# Reproducible Research: Peer Assessment 1
by **Jerzy Baranowski**

## Loading and preprocessing the data

```{r loading data}
data <- read.csv(unz("activity.zip", "activity.csv"))

data$interval<-as.character(data$interval)
for (i in 1:nrow(data))
{
        if(nchar(data$interval[i])<2) 
        {data$interval[i]<-paste("000",data$interval[i],sep="")}
        else if (nchar(data$interval[i])<3)
        {data$interval[i]<-paste("00",data$interval[i],sep="")}
        else if (nchar(data$interval[i])<4)
        {data$interval[i]<-paste("0",data$interval[i],sep="")}      
}
data$time<-paste(data$interval,data$date)
data$time<-strptime(data$time,"%H%M %F")
data$interval<-as.factor(data$interval)
data$date<-as.factor(data$date)
```

## What is mean total number of steps taken per day?
```{r mean computation}
library(plyr) #this package provides function ddply
# create summaries mean, median, sum for daily data with NA removed
daily<-ddply(data,.(date),# ddply creates a data frame from data splited by variable date
                summarise, # uses function summarise
                        #mean=mean(steps,na.rm=TRUE),
                        #median=median(steps,na.rm=TRUE),
                        sum=sum(steps,na.rm=TRUE)
              )
mean_steps<-mean(daily$sum)
median_steps<-median(daily$sum)
daily$mean<-mean(daily$sum)
daily$median<-median(daily$sum)
```
Using data from dataframe *daily* it is easy to compute:

- mean number of daily steps `r mean_steps`
- median of daily steps `r median_steps`



```{r histogram}
library(ggplot2)
ggplot(daily)+
        geom_histogram(aes(sum),binwidth=1000)+
        labs(title="Histogram of daily steps")


```


## What is the average daily activity pattern?
To compute daily activity pattern the following steps had to be taken

1. Create a new data frame
2. Convert the factor into hourly time
3. Convert hourly time int chron
4. Set GMT timezone (otherwise time ticks are shifted)
5. Create a plot
```{r daily pattern}

# create summaries mean, median, sum for interval data with NA removed
perminute<-ddply(data,.(interval),# ddply creates a data frame from data splited by variable date
             summarise, # uses function summarise
             median=median(steps,na.rm=TRUE), #will be used later
             mean=mean(steps,na.rm=TRUE)
)
#creating time variable from interval
perminute$time<-strptime(perminute$interval,"%H%M")
#converting time variable to chron, in order to plot it reasonably
library(chron)
perminute$time<-times(format(perminute$time, "%H:%M:%S"))
Sys.setenv(TZ='GMT')#necessary to avoid time shift
#ploting the time series of average steps per 5 minute interval
ggplot(perminute)+
        geom_line(aes(x=time,y=mean))+#line
        scale_x_chron(format="%H:%M")+#timescale
        labs(title="Average of steps per 5 minute intervals")+
        labs(y="Average steps")
```

## Inputing missing values

```{r missing data}
#count missing data
count_<-sum(!complete.cases(data$steps))
```
There are `r count_` missing step measurements. It is reasonable to replace them by median of appropriate time period as a missing value. The reasons are that it will always be an integer value and that in the intervals dominated by no activity it will be filled with zeros.
```{r filling missing data}
#find location of missing data
logical_index<-!complete.cases(data$steps)
#create a vector of factors
xt<-data$interval[logical_index]
#obtain appropriate medians
yt<-perminute$median[xt]
#create new data 
new_data<-data
#fill it with missing values
new_data$steps[logical_index]<-yt
```
## Are there differences in activity patterns between weekdays and weekends?
In order to verify the difference between weekdays and weekends a function is.weekend was created. 

```{r is.weekend}
is.weekend<- function(x){ 
        if (weekdays(x,abbreviate=T)=="sob" ||weekdays(x,abbreviate=T)=="nie")
        {
                return(TRUE)  
        }
        else {
                return(FALSE)
        }
}
```
Using this function the dataset was modified and stripped of unnecessary data
```{r}
for (i in 1:dim(new_data)[1])
{
        if (is.weekend(new_data$time[i])){
                new_data$weekend[i]<-"Weekend"
        }
        else{
                new_data$weekend[i]<-"Weekday"
        }
        
}

new_data$time<-NULL
new_data$date<-NULL
perminute2<-ddply(new_data,.(interval,weekend),summarise, mean=mean(steps))
perminute2$time<-strptime(perminute2$interval,"%H%M")
perminute2$time<-times(format(perminute2$time, "%H:%M:%S"))
perminute2$weekend<-as.factor(perminute2$weekend)

```
Using this data a panel plot is created

```{r Weekend}
ggplot(perminute2,aes(time,mean))+
        geom_line()+
        scale_x_chron(format="%H:%M")+
        facet_grid(weekend~.)+
        labs(title="Weekday vs weekend activity")


```

